---
title: "TMT10plex lisym data - timecourse"
author: "Nikola Muller"
date: "26.3.18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("~/planA/software tools utils/myUtils.R")
library(readxl)
library(pheatmap)
library(venn)
```

## Load data and extract expression

Load `proteinGroups_TMT.xls` file.
Use "Protein IDs" column as unique identifier or fows. And work with MaxQuant generated  `Reporter intensity corrected` values



```{r data_readin }

## dataset load + extract relevant info > expression file
data.input.file = read_xls("../data/proteinGroups_TMT.xls")

data.reporter.intens.corr = cbind(data.input.file[,c("Protein IDs","Majority protein IDs")], data.input.file[,grep("Reporter intensity corrected",colnames(data.input.file))])

data.reporter.intens.corr.expr = data.frame(lapply(data.reporter.intens.corr[,grep("TMT",colnames(data.reporter.intens.corr))],as.numeric))


rownames(data.reporter.intens.corr.expr) = unlist(data.input.file[,"Protein IDs"])

boxplot(log2(1+data.reporter.intens.corr.expr),las=2)

tmt.col.name = substring(colnames(data.reporter.intens.corr.expr),32,35)


```



### remove proteins where not measured in ALL pool reference channesl

Find reference channels with any 0 expression value and REMOVE them -> Essential cleaning!

```{r}

ref.channel.expr = sapply(unique(tmt.col.name),function(tmt_id){
	(data.reporter.intens.corr.expr[,paste("Reporter.intensity.corrected.0.",tmt_id,sep="")])
})

# mark those with all NA columns (why ever this happens I dont know)
mark2delete = which(apply(ref.channel.expr,MARGIN = 1, function(x)sum(is.na(x)))!=0)

zero.entry.in.refChannel = (apply(ref.channel.expr,MARGIN = 1, function(x)sum((x==0))) )
hist(zero.entry.in.refChannel)

mark2delete = unique(c(mark2delete,which(zero.entry.in.refChannel>0)))

length(mark2delete)



boxplot(log10(data.reporter.intens.corr.expr[,names(which(annotation.decoded[colnames(data.reporter.intens.corr.expr)] == "Pool"))]),main="MaxQuant corrected intensitites of all pooled samples",las=2)


pheatmap(cor(log10(ref.channel.expr[-1*mark2delete,]), use = "pairwise.complete"), main="Correlation of TMT reference (0) channel")




# boxplot(log2(1+data.reporter.intens.corr.expr),las=2)

tmt.col.name = substring(colnames(data.reporter.intens.corr.expr),32,35)



```


# I delete ``r length(mark2delete)`` proteins from expression matrix because 0 or NA values in REFERENCE channel!

```{r cleaning} 

data.reporter.intens.corr = data.reporter.intens.corr[-1*mark2delete,]


data.reporter.intens.corr.expr = data.reporter.intens.corr.expr[-1*mark2delete,]
dim(data.reporter.intens.corr.expr)


rm(data.reporter.intens.corr) ## to cleanup
rm(mark2delete)

```

That leaves ``r nrow(data.reporter.intens.corr.expr)`` realiably identified proteins






### load annotation info

Load annotation from matrix like file `sample.annotation.matrix.txt` and convert to tabular format

```{r read_annotation}






## add data descr 

annotation = read.table("../data/sample.annotation.matrix.txt",header=T,row.names="channel")
# colnames(data.reporter.intens.corr.expr)

annotation.decoded = c()
for(i in 0:9){
	for(j in paste("TMT",1:9,sep="")){
		annotation.decoded[paste("Reporter.intensity.corrected.",i,".",j,sep="")] = 
		  as.character(annotation[paste(i),j])
	}
}





annotation.decoded.tabular = do.call(rbind,lapply(strsplit(annotation.decoded,"_"),function(x){
	if(length(x)==3)
		{c(diet=(x[1]),time=sub("w","",x[2],ignore.case = T),mouse=x[3])}
	else
	{
		c(diet=paste(x[1],tolower(x[2]),sep="."),time=sub("w","",x[3],ignore.case = T),mouse=x[4])
	}
	
}))

annotation.decoded.tabular = cbind(annotation.decoded.tabular,"name" = annotation.decoded)
rownames(annotation.decoded.tabular) = annotation.decoded.tabular[,"name"]

head(annotation.decoded.tabular)

```




### compute ratios 
… as suggested from a maxquant discussion group
ratio_samplei = (1+expression_sample_i)/(1+matching_reference_channel_expression)

```{r}


sample.plex.ratios = do.call(cbind,lapply(unique(tmt.col.name),function(tmt_id){
	
	do.call(cbind,lapply(1:9,function(i){
		
		(1+data.reporter.intens.corr.expr[,paste("Reporter.intensity.corrected.",i,".",tmt_id,sep="")])/
		(1+data.reporter.intens.corr.expr[,paste("Reporter.intensity.corrected.0.",tmt_id,sep="")])
		
	}))
		
}))

colnames(sample.plex.ratios) = c(sapply(unique(tmt.col.name),function(x){sapply(1:9,function(y)paste("ratios.",y,".",x,sep=""))}))

rownames(sample.plex.ratios) = rownames(data.reporter.intens.corr.expr)

head(sample.plex.ratios)


rm(data.reporter.intens.corr.expr) ## clean


#remove samples as annotated by barbara and claire
# data.reporter.intens.corr.expr = data.reporter.intens.corr.expr[,!colnames(data.reporter.intens.corr.expr) %in% c("Reporter.intensity.corrected.6.TMT9","Reporter.intensity.corrected.6.TMT9","Reporter.intensity.corrected.6.TMT9","Reporter.intensity.corrected.6.TMT9")]
# 

sample.plex.ratios = sample.plex.ratios[,!colnames(sample.plex.ratios) %in% c("ratios.6.TMT9","ratios.6.TMT9","ratios.6.TMT9","ratios.6.TMT9")]

dim(sample.plex.ratios)


# sample order changed > new decoding
annotation.decoded.ratios = annotation.decoded
names(annotation.decoded.ratios) = sub("Reporter.intensity.corrected","ratios",names(annotation.decoded))
# remove reference channel
annotation.decoded.ratios = annotation.decoded.ratios[-1*grep("0",names(annotation.decoded.ratios))]




```



## inspect







```{r}

### now for ratios

annotation.data.frame = data.frame(cbind(annotation.decoded.tabular[annotation.decoded.ratios[colnames(sample.plex.ratios)],1:3],tmt=substr(colnames(sample.plex.ratios),10,13)),row.names = colnames(sample.plex.ratios))

pheatmap(cor(log10(sample.plex.ratios),use = "pair"),labels_row  = annotation.decoded.ratios[colnames(sample.plex.ratios)],labels_col = annotation.decoded.ratios[colnames(sample.plex.ratios)],main="Ratios",annotation_row =  annotation.data.frame)

# boxplot(log2(data.reporter.intens.corr.expr))
# boxplot(log10(sample.plex.ratios))




```





# plot time course & stats


define individual vectors with annotation content for easy use later on.
And define annotation & nice description thingy to use

```{r}


expression.annotation = annotation.data.frame[colnames(sample.plex.ratios),c("diet","time","mouse")]
expression.data.colnicename = annotation.decoded.ratios[colnames(sample.plex.ratios)]


time = as.numeric(expression.annotation[,"time"])
mouse.pch = c(15,16,17)[annotation.data.frame[,"mouse"]]
diet = as.character(annotation.data.frame[,"diet"])
diet.col = c("SD"="black","WD.4w"="blue","WD.6w"="orange","WD.8w"="red")[as.character(annotation.data.frame[,"diet"])]



```




```{r}


# boxplot(expression.data[,c(which(expression.data.colnicename %in% c("SD_12w_M1", "SD_6w_M1", "SD_30w_M2")))],las=2,main="presumably good?",ylim=c(-1,1))
# grid()

# boxplot(expression.data[,c(which(expression.data.colnicename %in% c("SD_34w_M1", "WD_6w_12W_M2", "WD_6w_20w_M2")))],las=2,,ylim=c(-1,1),main="presumably bad?")
# grid()


boxplot(log10(sample.plex.ratios[,order(expression.data.colnicename)]),las=2,,ylim=c(-1.5,1.5),main="overall ratios",col=diet.col[order(expression.data.colnicename)], outcol="lightgray")
grid()


```



Important: Especially for long fed diets the overall expression GOES DOWN

# So now we put values to same median



```{r}





expression.data.centered = apply(log10(sample.plex.ratios),MARGIN = 2, function(x)x/median(x,na.rm=T))
rownames(expression.data.centered) = rownames(sample.plex.ratios)

expression.data.scaled = apply(log10(sample.plex.ratios),MARGIN = 2, scale)
rownames(expression.data.scaled) = rownames(sample.plex.ratios)

boxplot(expression.data.centered[,order(expression.data.colnicename)],las=2,,ylim=c(-5,5),main="overall ratios centered only ",col=diet.col[order(expression.data.colnicename)], outcol="lightgray")
grid()

boxplot(expression.data.scaled[,order(expression.data.colnicename)],las=2,,ylim=c(-5,5),main="overall ratios scaled",col=diet.col[order(expression.data.colnicename)], outcol="lightgray")
grid()


pheatmap(cor((expression.data.centered),use = "pair"),labels_row  = annotation.decoded.ratios[colnames(sample.plex.ratios)],labels_col = annotation.decoded.ratios[colnames(sample.plex.ratios)],main="Ratios -- centered only",annotation_row =  annotation.data.frame, clustering_distance_rows = "euclidean",clustering_distance_cols = "euclidean")



pheatmap(cor((expression.data.scaled),use = "pair"),labels_row  = annotation.decoded.ratios[colnames(sample.plex.ratios)],labels_col = annotation.decoded.ratios[colnames(sample.plex.ratios)],main="Ratios -- scaled only",annotation_row =  annotation.data.frame,clustering_distance_rows = "euclidean",clustering_distance_cols = "euclidean")





```



Continue to work with scaled ratios. Look more biologically clustering.


```{r}




## LOCK IN: useing scaled expression for further analysis
expression.data = (expression.data.scaled)

# remove some objects
rm(expression.data.scaled, expression.data.centered,expression.annotation, expression.data.colnicename)

expression.annotation = annotation.data.frame[colnames(expression.data),c("diet","time","mouse")]
expression.data.colnicename = annotation.decoded.ratios[colnames(expression.data)]

dim(expression.data)

proteinID2geneName = tapply(unlist(data.input.file[,"Gene names"]),unlist(data.input.file[,"Protein IDs"]),c)


```


## do stats

Start simple: linear model accounting for time (but not time dependent) and diet
FDR correct to 0.01.

```{r}




table(time, diet)
time.ignoreinbetween = time
time.ignoreinbetween[ time %in% c(4, 6,7,9,10,11,12,13,14)] = NA
time.ignoreinbetween[time.ignoreinbetween == 2] = 1
# time.saveme = time
time = time.ignoreinbetween

## rough stats
diet_overall = apply(expression.data,MARGIN = 1,function(p){
	lm(p ~ time.ignoreinbetween + diet)
})

assertthat::assert_that(length(diet_overall)==nrow(expression.data))


# p.values = unlist(lapply(diet_overall,function(i){
# 	fstat = summary(i)$fstatistic
# 	pf(fstat[1], fstat[2], fstat[3], lower.tail=FALSE) 
# }))
# 
# fdr = p.adjust(p.values,method = "BH")
# sum(p.values < 0.01)
# sum(fdr < 0.01)
# which.min(fdr)




p.values.byCovariate = do.call(rbind,lapply(diet_overall,function(i){
	coefficients(summary(i))[-1,"Pr(>|t|)"]
}))
rownames(p.values.byCovariate) = rownames(expression.data)


assertthat::assert_that(!is.null(rownames(p.values.byCovariate)))

p.values.byCovariate.fdr = apply(p.values.byCovariate,MARGIN = 2, p.adjust,method="BH")



apply(p.values.byCovariate.fdr,2,function(i)sum(i<0.01))


venn(apply(p.values.byCovariate.fdr[,-1],2,function(i){which(i<0.01)}))


```



```{r}


plotSingleProteinExpr = function(id){
	### PLOT sample expression
y = expression.data[id,]

	plot(time,y,col=diet.col,pch=mouse.pch,type="p",main=paste(id,proteinID2geneName[id]))


}



```

```{r}

# WD 4 weeks proteins:
top_4weeks = expression.data[names(which(p.values.byCovariate.fdr[,"dietWD.4w"]<0.001)),]
pheatmap(top_4weeks[,order(expression.data.colnicename)],cluster_cols = F,annotation_col = expression.annotation,main="FDR 0.001 proteins in WD 4w",labels_col = expression.data.colnicename[order(expression.data.colnicename)],labels_row = substring(rownames(top_4weeks),1,10))

pheatmap(top_4weeks[,order(expression.data.colnicename)],cluster_cols = F,annotation_col = expression.annotation,main="FDR 0.001 proteins in WD 4w",labels_col = expression.data.colnicename[order(expression.data.colnicename)],labels_row = proteinID2geneName[rownames(top_4weeks)])


# interesting guys are: P55050 = Fatty acid-binding protein, intestinal
# Q921H8  = 3-ketoacyl-CoA thiolase A, peroxisomalGene: Acaa1a
# Q8VCH0  = 3-ketoacyl-CoA thiolase B, peroxisomal Gene: Acaa1b
# P84244;P02301 = Histone H3.3 (Gene: H3f3a) & Histone H3.3C (Gene: H3f3c)

## plot

# plotSingleProteinExpr("P55050")
# plotSingleProteinExpr("Q921H8")
# plotSingleProteinExpr("Q8VCH0")
# plotSingleProteinExpr("P84244;P02301")

plotSingleProteinExpr("P54797")
plotSingleProteinExpr("P62806")



```


```{r}

# WD 6 weeks proteins:
top_6weeks <- expression.data[names(which(p.values.byCovariate.fdr[,"dietWD.6w"]<0.000001)),]
pheatmap(top_6weeks[,order(expression.data.colnicename)],cluster_cols = F, annotation_col = expression.annotation,main="FDR 0.000001 proteins in WD 6w",labels_row = substring(rownames(top_6weeks),1,20),fontsize = 6,labels_col = expression.data.colnicename[order(expression.data.colnicename)])


proteinID2geneName[rownames(top_6weeks)]

plotSingleProteinExpr("Q920E5")


```



```{r}


# WD 8 weeks proteins:
top_8weeks <- expression.data[names(which(p.values.byCovariate.fdr[,"dietWD.8w"]<0.0000001)),]
pheatmap(top_8weeks[,order(expression.data.colnicename)],cluster_cols = F,annotation_col = expression.annotation,main="FDR 0.0000001 proteins in WD 8w",labels_row = substring(rownames(top_8weeks),1,20),fontsize = 6,labels_col = expression.data.colnicename[order(expression.data.colnicename)])




proteinID2geneName[rownames(top_8weeks)]

plotSingleProteinExpr("Q920E5")



```


```{r}

# WD 4 weeks proteins:
only_4weeks_sign = expression.data[names(which(p.values.byCovariate.fdr[,"dietWD.4w"]<0.01 & p.values.byCovariate.fdr[,"dietWD.6w"]>0.01 & p.values.byCovariate.fdr[,"dietWD.8w"]>0.01)),]
pheatmap(only_4weeks_sign[,order(expression.data.colnicename)],cluster_cols = F,annotation_col = expression.annotation,main="only sign in WD 4w",labels_col = expression.data.colnicename[order(expression.data.colnicename)],labels_row = substring(rownames(top_4weeks),1,10))

pheatmap(only_4weeks_sign[,order(expression.data.colnicename)],cluster_cols = F,annotation_col = expression.annotation,main="only sign in WD 4w",labels_col = expression.data.colnicename[order(expression.data.colnicename)],labels_row = proteinID2geneName[rownames(top_4weeks)])


# interesting guys are: P55050 = Fatty acid-binding protein, intestinal
# Q921H8  = 3-ketoacyl-CoA thiolase A, peroxisomalGene: Acaa1a
# Q8VCH0  = 3-ketoacyl-CoA thiolase B, peroxisomal Gene: Acaa1b
# P84244;P02301 = Histone H3.3 (Gene: H3f3a) & Histone H3.3C (Gene: H3f3c)

## plot
plotSingleProteinExpr("P17717")
P24472

```

